Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland

### 2017-08-22 paper/main.tex
added comments from Dave

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used various markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we herein use single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that only few clusters could be fully characterized by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
\newpage

\section{Introduction}
The purpose of the mammary gland is to provide nourishment and passive immunity to the young until they are capable of feeding themselves.
From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}
In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fatpad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}
At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fatpad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
Towards the end of pregnancy the gland is extremely dense and primarily occupied by epithelial cells and very little fat.
This morphology is largely maintained throughout lactation.
However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006} .
Towards the end of involution the gland reaches a morphology resembling that prior to pregnancy and subsequent pregnancies will trigger the same chain of events.

Recent efforts have focused on the identification and characterisation of the various mammary epithelial cell lineages within the gland that contribute to this developmental homeostasis.
Pioneering fat-pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and shown that the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in the luminal cells\autocite{Asselin2007,Kouros2006}.
More recently, elegant lineage-tracing studies used key markers to address the contribution of each lineage to adult mammary epithelial cell homeostasis\autocite{Inman2015}.
However, in all of these studies only a handful of markers and genes were used to define the cellular hierarchy of the mammary epithelial cells, with a principal focus on the nulliparous developmental stage.
Therefore, to properly understand its changing role throughout life, there is a need for an unbiased and comprehensive characterisation of mammary epithelial cell compartments at different developmental stages.

To this end we herein use single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.

@@

First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI) (Supplementary Figure 2).
For each time point we sorted mammary epithelial cells based on the Epcam cell surface marker from two independent mice (\autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
After removal of enodthelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters). 
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).

@@

We would like to thank the staff at Sanger Institute, Research Service Facility (RSF) for their assistance.
We would like to thank Dr. Aaron T. Lun (CRUK CI) for helpful discussions and comments on manuscript.
K.B. is funded by a Cambridge Cancer Centre studentship.
D.A. is funded by CRUK and Wellcome Trust.
S.P. is funded by CRUK.
J.C.M is funded by CRUK and EMBL.
W.T.K is funded by a CRUK career establishment award (C47525/A17348), University of Cambridge and Magdalene College, Cambridge.

>>>>

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used various markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we herein used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that only a few clusters could be fully characterized by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
\newpage

\section{Introduction}
The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}
In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fatpad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}
At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fatpad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
Towards the end of pregnancy the gland is extremely dense and primarily occupied by epithelial cells and very little fat.
This morphology is largely maintained throughout lactation.
However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006} .
Towards the end of involution the gland reaches a morphology resembling that of glands prior to pregnancy and subsequent pregnancies will trigger the same chain of events.

Recent efforts have focused on the identification and characterisation of the various mammary epithelial cell lineages within the gland that contribute to this developmental homeostasis.
Pioneering fat-pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and showed the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in luminal cells\autocite{Asselin2007,Kouros2006}.
More recently, elegant lineage-tracing studies used key markers to address the contribution of each lineage to adult mammary epithelial cell homeostasis\autocite{Inman2015}.
However, in all of these studies only a handful of markers and genes were used to define the cellular hierarchy of the mammary epithelial cells, with a principal focus on the nulliparous developmental stage.
Therefore, to properly understand its changing role throughout life, there is a need for an unbiased and comprehensive characterisation of mammary epithelial cell compartments at different developmental stages.

Here we used single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.

@@

First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI) (Supplementary Figure 2).
For each time point we sorted mammary epithelial cells based on the Epcam cell surface marker from two independent mice (\autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters). 
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).

@@

We would like to thank the staff at Sanger Institute, Research Service Facility (RSF) for their assistance.
We would like to thank Dr. Aaron T. Lun (CRUK CI) for helpful discussions and comments on the manuscript.
K.B. is funded by a Cambridge Cancer Centre studentship.
D.A. is funded by CRUK and Wellcome Trust.
S.P. is funded by CRUK.
J.C.M is funded by CRUK and EMBL.
W.T.K is funded by a CRUK career establishment award (C47525/A17348), University of Cambridge and Magdalene College, Cambridge.
